The Galactic Midplane is Not a Plane: Implications for {\em Gaia}

### 2019-04-08 paper/ms.tex
new title
title & abstract edits

\title{The Midplane is Not a Plane: Implications for {\em Gaia}}

@@

Stellar actions, computed for stars from their 6D phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and are thus
good labels of a star's orbit. Of course, inaccurate measurements of the 6D
phase space position and uncertainities in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the 6D phase space position induces a
phase-dependence in the actions, which we interpret as an error. This error is
non-Gaussian and bimodal, with neither of the modes peaking on the true action
value. A midplane error of $\sim30\,\pc$ and $\sim225\,\pc$ is necessary to
induce a $100\,\%$ error in the vertical action $J_z$ for thin and thick disk
orbits, respectively. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic offset in $z$. Variation in the local standard of rest
will induce similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across large regions of the disk
accessible with the {\em Gaia} mission.

@@

\section{Introduction} \label{sec:intro}
Our understanding of the Milky Way is currently undergoing a revolution in the
era of {\em Gaia} DR2. Recent major discoveries include the remnants of a
major merger
\citep{2018ApJ...860L..11K,2018Natur.563...85H,2018arXiv180704290L,2019MNRAS.482.3426M},
a phase-space ``spiral'' in the solar neighborhood \citep{2018Natur.561..360A} possibly indicating either local
substructure infall \citep{2018MNRAS.481.1501B,2018arXiv180800451L} or bar
buckling \citep{2018arXiv181109205K}, and a gap suggestive of perturbation by a dark matter
subtructure in the tidal stream GD1 \citep{2018arXiv181103631B}.
These discoveries all indicate that the Milky Way's stellar distribution, which demonstrably departs significantly from axisymmetry, is undergoing phase mixing and dynamical interactions across a range of spatial and temporal scales.

@@

Once a Galactocentric coordinate system has been established and a 6D phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities which describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the typical integral of the Hamiltonian in cylindrical coordinates and are
given by,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,}
\eeq
where $i=R,z,\phi$ are the cylindrical radial, vertical, and azimuthal coordinates respectively and the $p_i$ are the conjugate momenta.\footnote{We typically write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving
axisymmetric potential, these actions are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.

With the advent of 6D phase-space measurements over a relatively large ($\gtrsim 2$ kpc) volume from the {\em Gaia} satellite, the study of stars' actions has gained new popularity. One reason is dimensionality reduction---an individual stellar orbit is concisely described by three actions, as opposed to six dimensions of phase space. Under the assumption of a phase-mixed system, the dynamical properties of a population of stars should be uniquely a function of their actions and independent of the conjugate angles, allowing investigation of the
relationship between {\em orbital} properties of stars and other intrinsic, and at least partially invariant, properties such as age or metallicity. Actions also provide a convenient basis for constructing models of the stellar distribution function, or for associating stars with similar dynamical properties (e.g. to determine membership in moving groups).

If the system being considered departs from axisymmetry in a significant and non-adiabatic way, the actions computed using an axisymmetric approximation to the true potential can exhibit cyclic dependence on the orbital phase (or time at which the star's position and velocity are observed), large-scale migration, or diffusion from their initial values.
In the Milky Way, stellar actions are expected to diffuse on short time scales due to scattering with gas clouds
and to evolve on longer time scales in the case of orbits near resonances with spiral arms, bar(s), and other
large scale perturbations \citep{2014RvMP...86....1S}. For this reason, we and
other authors have used actions to study stellar scattering in the Milky Way
disk using the improved astrometry of {\em Gaia} DR2 and various age
catalogues \citep{2018ApJ...867...31B,2018arXiv180803278T}. Actions have also
been used to study different models of spiral structure in the Milky Way
\citep{2019MNRAS.tmp..155S}. Characteristics of the distribution of stars in the extended Solar neighborhood in action space are discussed in
\citet{2018arXiv180503653T}.

The transformation to action space is nonlinear in both the phase-space position and the potential, and requires relatively strong assumptions about the system's symmetry. For instance, the 
Galactic potential is not strictly axisymmetric; with the vast improvement in the quality of phase-space measurements due to {\em Gaia}, this assumption appears increasingly inadequate \citep[e.g.][]{2018Natur.561..360A}. Even if this assumption is close enough for many purposes, the parameters used in axisymmetric models may be inadequately constrained by current observations.

%In addition, the necessity of 6D phase space measurements of a star limits the Galactic volume observable in action space.

%With these limitations in mind, it is therefore natural to consider the reason to transform to action space.

In this work we discuss how the imposition of a global axisymmetric coordinate system based on a local measurement of the midplane location and Galactocentric distance introduces systematic error in the computation of actions, especially for stars in the thin disk. Our main point
is that the ``local'' Galactic midplane is not the same as the ``global''
Galactic midplane. For stars relatively far from the Sun,
the local midplane location will not agree with
our local midplane extrapolated onto their position, thanks to relatively small-scale variations in the density of the gas and stellar distribution, not necessarily even as large as spiral arms. Converting positions and velocities of more distant stars from a heliocentric to a
Galactocentric coordinate system introduces a systematic bias in the $z$ coordinate relative to $z$ velocity, and hence the actions computed for these stars. The further from the solar neighborhood the target star, the more likely the mismatch will result in large systematic uncertainty, especially in the vertical action $J_z$. Depending on the type of orbit being studied, this sets a limit on the ``dynamical solar neighborhood'' over which the local coordinate frame can be usefully extended for action-space analysis. A similar argument applies to uncertainty in measurements of
$R_{\odot}$ and variations in the local standard of rest (LSR). In Section~\ref{sec:ref_frame}, we describe the general impact coordinate
system errors have on the measured actions.

Lacking an empirical, nonlocal definition of the global Galactic midplane or a measurement of the expected midplane variation in the Milky Way itself,  in Section~\ref{sec:local_fire} we attempt to quantify the azimuthal dependence of the
local midplane, and hence the expected size of the dynamical solar neighborhood, using cosmological-hydrodynamical zoom-in simulations of Milky Way-mass galaxies from the Feedback in Realistic Environments (FIRE)
collaboration\footnote{\url{https://fire.northwestern.edu}} \citep{2016ApJ...827L..23W,2017arXiv170206148H}.

In Section~\ref{sec:discussion} we discuss the implications of midplane variations, and the resulting systematic uncertainty in the vertical action, for action-space analyses beyond the dynamical solar neighborhood. As an illustration, in Section \ref{sssec:neighborhood} we use these estimates to consider the feasibility of one science goal that relies on the invariance of actions --- the reconstruction of open clusters across large distances using ``dynamical tagging;'' that is, the association of stars to a birth cluster based on their similar actions. In light of our findings, we considered the orbits of the
nine open clusters within $250\,\pc$ as reported by the {\em Gaia}
collaboration \citep{2018AA...616A..10G}. In Table~\ref{tab:real_clusters}, we
report the distance to each cluster, the three actions, and the maximum
vertical extent of their orbits ($z_{\text{max}}$). 
%The actions are computed using the technique described in Section~\ref{ssec:action_comp}. 
 In Section~\ref{sssec:reconstruction} we estimate and discuss the size of the dynamical solar neighborhood for each cluster based on its mass, to show that action-space tagging is likely still a viable method if background is unimportant.  We summarize and conclude in Section~\ref{sec:conclusion}.

@@

@@

\section{Conclusions}\label{sec:conclusion}
Actions have promise as excellent orbit labels. If the Galaxy can be
approximated as axisymmetric and 6D phase space positions can be measured
accurately and precisely enough, then the computed actions are invariant with
orbital phase. However, we have shown that the fact that the Galactic midplane
is not constant across the disk presents a significant complication to
computed actions actually being invariant. Our main conclusions are as
follows:

@@

Add sentences about distribution function modeling

Our main point is that the local midplane varies between different points in
the Galaxy. This is a different statement than that our measurements of the
local midplane are erroneous. Indeed, the local midplane has recently been
measured to a claimed high level of precision, setting $z_{\odot} = 20.8 \pm
0.3\,\pc$ \citep{2019MNRAS.482.1417B}. However, we also do not mean to say
that determinations of the local midplane are perfect, since midplane
measurements from H~\textsc{ii} regions place $z_{\odot} \sim 5-10\,\pc$. An
additional complication is the fact that non-zero Solar heights induce a tilt
in the Galactic midplane \citep{2014ApJ...797...53G,2016ARAA..54..529B}. We do not propose here what a modeler should use for the ``true'' global midplane, since it depends on the particular problem being studied. Even without a high-precision definition of the global Galactic midplane, current observations from Gaia should soon permit a measurement of the real azimuthal dependence of the midplane location.

>>>>

\title{The Galactic Midplane is Not a Plane: Implications for {\em Gaia}}

@@

Stellar actions, computed for stars from their six-dimensional phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. Of course, inaccurate measurements of the 
phase space position and uncertainities in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as an error. This error is
non-Gaussian and bimodal, 
%mm [this sounds imprecise: an error wouldn't peak on a value...]
with neither of the modes peaking on the true action value. 
A midplane error of $\sim30\,\pc$ for a thin disk orbit or
$\sim225\,\pc$ for a thick disk orbit is all that is necessary to
induce a $100\,\%$ error in the vertical action $J_z$.
orbits, respectively. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic offset in $z$. Variation in the local standard of rest
will induce similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across large regions of the disk
accessible with the {\em Gaia} mission.

@@

\section{Introduction} \label{sec:intro}
Our understanding of the Milky Way is currently undergoing a revolution in the
era of {\em Gaia} DR2. Recent major discoveries include the remnants of a
major merger \citep{2018ApJ...860L..11K, 2018Natur.563...85H,
2018arXiv180704290L, 2019MNRAS.482.3426M}, a phase-space ``spiral'' in the
solar neighborhood \citep{2018Natur.561..360A} possibly indicating either
local substructure infall \citep{2018MNRAS.481.1501B,2018arXiv180800451L}, bar
buckling \citep{2018arXiv181109205K}, although may just be contributions from
moving groups \citep{2019arXiv190308325M} and a gap suggestive of perturbation
by a dark matter subtructure in the tidal stream GD1
\citep{2018arXiv181103631B}. These discoveries all indicate that the Milky
Way's stellar distribution, which demonstrably departs significantly from
axisymmetry, is undergoing phase mixing and dynamical interactions across a
range of spatial and temporal scales.

@@

Once a Galactocentric coordinate system has been established and a 6D phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities which describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the typical integral of the Hamiltonian in cylindrical coordinates and are
given by,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,}
\eeq
where $i=R,z,\phi$.\footnote{We typically write $J_r$ instead of $J_R$ for
aesthetic reasons.} Note that $J_{\phi} \equiv L_z$. In a slowly-evolving
axisymmetric potential, the actions of stars' orbits are conserved quantities
\citep{2008gady.book.....B,2014RvMP...86....1S}. The non-axisymmetries of a
more realistic potential induce time evolution in the {\em measured} actions
--- the true orbital invariants (if they exist) are invariant by definition.

In a slowly-evolving, axisymmetric potential, the actions of each star are
time-invariant \citep{2008gady.book.....B}. In the Milky Way, stellar actions
are expected to evolve on short time scales due to scattering with gas clouds
and on long time scales due to resonances with spiral arms, bar(s), and other
large scale perturbations \citep{2014RvMP...86....1S}. For this reason, we and
other authors have used actions to study stellar scattering in the Milky Way
disk using the improved astrometry of {\em Gaia} DR2 and various age
catalogues \citep{2018ApJ...867...31B,2018arXiv180803278T}. Actions have also
been used to study different models of spiral structure in the Milky Way
\citep{2019MNRAS.tmp..155S}. A summary of the disk in action space is given by
\citet{2018arXiv180503653T}.

The transformation to action space is not trivial, and many of the assumptions
of the transformation are potentially incorrect. For instance, the assumption
that the Galactic potential is axisymmetric may be inadequate with current
measurement errors. Even if this assumption is valid, the parameters used in
our axisymmetric models may be inadequately tuned to observations. In
addition, the necessity of 6D phase space measurements of a star limits the
Galactic volume observable in action space.

With these limitations in mind, it is therefore natural to consider the reason
to transform to action space. One reason is dimensionality reduction --- in
the action space paradigm, a stellar orbit is completely described by three
actions, as opposed to six dimensions of phase space. This means that the
relationship between {\em orbital} properties of individual stars and stellar
properties such as age or metallicity can be investigated.

We explore one science goal that relies on the conservation of actions --- the
reconstruction of open clusters across large distances using dynamical
tagging. The basic program is to match stars with roughly the same actions as
their supposed natal cluster. To set the scene, we consider the orbits of the
nine open clusters within $250\,\pc$ as reported by the {\em Gaia}
collaboration \citep{2018AA...616A..10G}. In Table~\ref{tab:real_clusters}, we
report the distance to each cluster, the three actions, and the maximum
vertical extent of their orbits ($z_{\text{max}}$). The actions are computed
using the technique described in Section~\ref{ssec:action_comp}. We also show
the mass of each cluster and the ``neighborhood'' we compute for each cluster,
which gives a rough sense of how far one can move away from the cluster before
the issue which we discuss next becomes important. This neighborhood is
discussed in more detail in Section~\ref{sssec:reconstruction}.

@@

We propose a new type of error in the computation of actions which has
implications for general dynamical modelling across the disk. Our main point
is that the ``local'' Galactic midplane is not the same as the ``global''
Galactic midplane. If we observe a group of stars that are
non-local\footnote{i.e. a group of stars that is far enough away. The exact
distance depends on the accuracy of the measurement being made and the goals
of the particular problem.}, they will know of a different local midplane than
our local midplane extrapolated onto their position. This means that when we
convert those star's positions and velocities from a heliocentric to a
Galactocentric coordinate system, we are inducing a systematic bias in those
stars' $z$ values. The same applies to inaccuracies in measurements of
$R_{\odot}$ and variations in the local standard of rest (LSR).

We do not propose what a modeller should use for the ``true'' global midplane,
since such a definition probably depends on the particular problem such a
modeller is facing. Even without a high-precision definition and measurement
of the global Galactic midplane, we can still assess how different we should
expect the local midplane to be by measuring the azimuthal dependence of the
local midplane. This procedure may be readily tractable over some of the disk
with current datasets, using appropriate selection functions. However, we turn
to simulation in this work because we can investigate the entire disk.
Specifically, we use cosmological zoom-ins of Milky Way-mass galaxies from the
Feedback in Realistic Environments (FIRE)
collaboration\footnote{\url{https://fire.northwestern.edu}} (the {\em Latte}
suite of simulations).

Our main point is that the local midplane varies between different points in
the Galaxy. This is a different statement than that our measurements of the
local midplane are erroneous. Indeed, the local midplane has recently been
measured to a claimed high level of precision, setting $z_{\odot} = 20.8 \pm
0.3\,\pc$ \citep{2019MNRAS.482.1417B}. However, we also do not mean to say
that determinations of the local midplane are perfect, since midplane
measurements from H~\textsc{ii} regions place $z_{\odot} \sim 5-10\,\pc$. An
additional complication is the fact that non-zero Solar heights induce a tilt
in the Galactic midplane \citep{2014ApJ...797...53G,2016ARAA..54..529B}.

In Section~\ref{sec:ref_frame}, we describe the general impact coordinate
system errors have on the measured actions. In Section~\ref{sec:local_fire},
we measure the local midplane in cosmological zoom-in simulations from the
FIRE collaboration. In Section~\ref{sec:discussion} we discuss and give some
context for our results, and implications for the definition of the Solar
neighborhood. We conclude in Section~\ref{sec:conclusion}.

@@

\section{Conclusions}\label{sec:conclusion}
Actions have promise as excellent orbit labels. If the Galaxy can be
approximated as axisymmetric and 6D phase space positions can be measured
accurately and precisely enough, then the computed actions are invariant with
orbital phase. However, we hvae shown that the fact that the Galactic midplane
is not constant across the disk presents a significant complication to
computed actions actually being invariant. Our main conclusions are as
follows:

@@
